Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-01-19 LICS/sections/introduction.tex
wip

\ifdefined \ONECOLUMN
\begin{figure}
	\[
		\scalebox{0.55}{
		\tikzfig{../figures/categorical-semantics/egraph-translation}
		}
	\]
	\caption{Example translation of acyclic e-graphs into string diagrams for semilattice enriched Cartesian categories. }
	\label{fig:e-graph-example}
	\end{figure}
\else
\begin{figure*}
\[
	\hspace{1.3cm}
    \scalebox{0.65}{
    \tikzfig{../figures/categorical-semantics/egraph-translation}
    }
\]
\caption{Example translation of acyclic e-graphs into string diagrams for semilattice enriched Cartesian categories. }
\label{fig:e-graph-example}
\end{figure*}
\fi

@@

In order to \textit{implement} generalised e-graphs, rewriting must be performed on concrete representations of the corresponding string diagrams.  
String diagrams can be considered equivalently as either topological objects (\textit{i.e.}, taken modulo ``connectivity'') or as a 2-dimensional syntax quotiented by the equations of a \textit{symmetric monoidal category (SMC)}.
For example, we have the following equivalences between diagrams:
%\[(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)\] 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/interchange}
	}
\]
The latter representation makes string diagrams unamenable to efficient implementation due to the difficulty of calculating the quotient. 
A large body of research work shows how this issue can be solved by taking a different approach, representing string diagrams as hypergraphs~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.  Here, the generators $c_i$ become hyper-edges,  and wires become vertices.  Thus the expected quotient becomes simply hypergraph isomorphism. With appropriate restrictions on the form these hypergraphs can take, this approach can be used to \textit{characterise} the free symmetric monoidal category generated by some~$c_i$.

We are interested not only in the free SMC over a set of generators, but also in \textit{(symmetric) monoidal theories} with extra equations such as, for instance, the rewrite rules (b)--(d) of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus rewrites of their hypergraph representations.  Because the generating equations can be applied in any context, we are led to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ within a hypergraph and replace it with (a sub-hypergraph corresponding to) $r$.
The standard methodology to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting} \cite{dpo, bonchi_string_2022} is still applicable.
However, these concepts must be now generalised from hypergraphs to \textit{e-hypergraphs}, hypergraphs with two additional relations denoting the hierarchical structure introduced by so-called \textit{e-boxes}: the generator for semilattice enrichment, and the separation of the components of each e-box.
Giving the appropriate definitions turns out to be a technical matter of some complexity, and constitutes the main body of the paper.

\subsection{Soundness and Completeness}
The main technical result of the paper is a proof of soundness and completeness of this representation.  This extends the results of \cite{bonchi_string_2022-2} for plain SMCs. While the structural equalities of SMCs are factored out in the representation,  those arising from the enrichment (see the equations of Figure \ref{fig:string-equations}) are not, and should not be. They represent the un/sharing of subdiagrams with respect to the e-box structure,  which is precisely what allows for the compact representation of equivalence classes.  Instead, we consider DPO-rewrites implementing both the structural equalities for enrichment (which involve the e-box structure) and the equalities arising from the generating monoidal theory (which do not).

Precisely,  our soundness and completeness result is the statement that morphisms in an appropriate free semilattice-enriched SMC are equal  if and only if there exists a sequence of DPO-rewrites --- each induced by a structural equality or the monoidal theory --- between their combinatorial representations. 
In the particular case of a monoidal signature including natural copy and delete maps,  we thus recover a sound and complete representation of semilattice-enriched Cartesian categories,  with the translation described earlier in the introduction justifying our claim of developing a mathematical theory of e-graphs.

%\subsection{E-matching and E-rewriting}
%
%Having defined e-hypergraphs and shown them to correctly model the rewriting of string diagrams for semilattice enriched SMCs, we note that the naive implementation of DPO-rewriting is inefficient: to find a redex within an e-hypergraph involves finding a structurally equivalent e-hypergraph which contains the redex as a subgraph. In general, this involves unsharing the e-box structure before a redex becomes available.  Thus, we define a notion of \textit{e-matching} for e-hypergraphs: that is, to find redexes working modulo the e-box structure. In particular, we wish to locate the smallest subgraph $G' \subseteq G$ "containing" (in an appropriate sense) the redex $L$. Given this,  \textit{e-rewriting} $L \to R$, intended to achieve equality saturation, is simple to define due to its non-destructive nature: we rewrite $G' \to G' + R$ in $G$.  
%\begin{figure}
%\[
%	\tikzfig{combinatorial_semantics/example-rewriting}
%\]
%\caption{Example application of e-rewriting for rewrite $(x*y)/z \to x*(y/z)$: find an appropriate minimal sub-e-hypergraph containing the redex $(x*y)/z$ and non-destructively add the reduct $x*(y/z)$.  REDO THIS}
%\end{figure}

@@

% E-graphs
Although e-graphs were first developed in the 80s \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimisers and synthesisers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries (``relational e-matching'') \cite{zhang_relational_2022}.  We are hopeful that our novel perspective on e-graphs can be extended to give uniform foundations to these investigations.  Some further avenues for investigation, including potential applications, are given in the conclusion of this paper.

@@

\section{Categorical Semantics of E-Graphs}
In this section,  we introduce the required preliminaries on semilattice enriched symmetric monoidal categories generated by monoidal theories,  and the string diagrammatic formalism we will use to represent them.

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  We write the identity morphism on $A$ as $\id_A$,  or simply $A$.  We commonly write $f;g$ for composition in diagram order.  Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$ and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.  
We denote by $\catname{SLatt}$ the closed monoidal category of semilattices which is defined in the Appendix (\textcolor{red}{add definition of $\catname{SLatt}$ in the Appendix}).

@@

It will often be convenient to give presentations of SMTs in terms of string diagrams.
\begin{example}
The SMT of \textit{commutative comonoids} is given by the following generators, ${\Delta, !}$, depicted in string diagrammatic form:
\[
	\scalebox{0.8}{
  	 \tikzfig{../figures/categorical-semantics/Cartesian-equipment}
	}
\]
together with the following associativity, commutativity and unitality equations, again given in terms of string diagrams. 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-theory}	
	}
\]
A Cartesian SMT is given by a set of generators $\Sigma_C$ which additionally contains the generators of commutative comonoids, together with the equations of the SMT of commutative comonoids, and the following additional naturality equations, for every $c \in \Sigma_C$
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-naturality}
	}
\]
By Fox's Theorem~\cite{fox},  $\textbf{S}(\Sigma_C, \mathcal{E}_C)$ is equivalent to the free Cartesian category over $\Sigma_C$. 
\end{example}

@@

We will use enrichment to axiomatize the e-box (e-class) structure as the ``join" of two morphisms $f,g: A \to B$ as $f + g: A \to B$. 
More precisely our hom-sets will be given the structure of a semilattice.
% Note, in particular,  that we do not require semilattices to have a unit for $+$. 
% This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must be a categorical biproduct \cite{maclane}---a degeneracy we wish to avoid.  
Generally,  we may define a category \textit{enriched} in any other (closed) monoidal category $M$ (\textit{e.g.},  \textbf{SLatt}).  
In this case,  hom-sets are generalised to hom-objects of $M$ (\textit{e.g.},  a semilattice) and composition 
\[
	\circ: Hom(B,C) \otimes Hom(A,B) \to Hom(A,C)
\]
is defined as an $M$-morphism which satisfies certain axioms.
Composition being an $M$-morphism implies it respects the additional structure on hom-sets.
%  In other words,  composition must respect the additional structure on hom-sets.  
% In our case,  the category we wish to enrich is also monoidal,  so we additionally ask that the monoidal structure also respects this structure.  Technically, this amounts to asking that the tensor product is an enriched functor. 
We omit further details of the general case of enrichment here, and instead work with the concrete definition required for our application, where $M = \textbf{SLatt}$.
It remains to be seen if more general enrichments have interesting applications.

\begin{definition}[Semilattice-enriched category]
A semilattice-enriched category $\mathbb{C}$ is defined by the following data
\begin{itemize}
	\item a set of objects $\mathbb{C}$
	\item for every pair of objects $A,B \in \mathbb{C})$ --- a hom-object $\mathbb{C}(A,B \in \catname{SLatt}$
	\item for every triple of objects $A,B,C \in \mathbb{C}$ --- a composition morphism
	\[
		\circ: \mathbb{C}(B,C) \otimes \mathbb{C}(A,B) \to \mathbb{C}(A,C)
	\]
	\item for every object $A \in C$ --- a unit morphism
	\[
	u_{A} : I \to \mathbb{C}(A,A), \text{ where $I$ is the monoidal unit for $\catname{SLatt}$}
	\]
\end{itemize}
such that particular coherence diagrams commute~\cite{Borceux_1994}.
\end{definition}

@@

Note, in particular,  that we do not require semilattices to have a unit for $+$, \textit{i.e.,} we work with \textit{unbounded} semilattices.
We also will not utilize the ordering that is induced by the operator of a semilattice in any meaningful way, so a semilattice can be thought of as a commutative idempotent semigroup throughout the paper.
We stick to the term semilattice as it is shorter.

\begin{definition}[Semilattice-enriched SMC]\label{def:enriched-prop}
A \textit{semilattice enriched strict SMC}  $\mathbb{C}$ is given by a semilattice-enriched category $\mathbb{C}$ as above that additionally has
\begin{itemize}
\item a unit object $I_{C} \in \mathbb{C}$
\item for every pair of objects $A,B \in \mathbb{C}$ --- an object $A \otimes B \in \mathbb{C}$
\item for all $A,B,C,D$ --- a tensor morphism $ - \otimes_{C} - : \mathbb{C}(A,C) \otimes \mathbb{C}(B,D) \to \mathbb{C}(A \otimes B, C \otimes D)$
\end{itemize}
such that particular coherence diagrams commute~\cite{enriched_monoidal}.
\end{definition}
The latter morphism being in $\catname{SLatt}$ implies the following equations that $\otimes$ satisfy
\[
f \otimes (g+h) = f \otimes g + f \otimes h \qquad (f+g) \otimes h = f \otimes h + g \otimes h
\]

@@

\begin{proposition}(A specialized case of Proposition 6.4.7~\cite{Borceux_1994})
There is a 2-adjunction 
% https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
\[\begin{tikzcd}
	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
	\end{tikzcd}
\]
that is induced by a usual free-forgetful adjunction

@@

In particular, 2-functor $F$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLatt}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
% SMC $(\mathbb{C}, \otimes, 1,+)$ where every hom-set additionally has the structure of a semilattice $(\mathbb{C}(A,B), +)$ -- that is,  a set equipped with an associative,  commutative,  and idempotent operation -- which respects the composition and tensor product in the following ways,  for all appropriately typed morphisms.
% \begin{align*}
% f ; (g+h) &= f;g + f;h &
% (f+g) ; h &= f;h + g;h \\
% f \otimes (g+h) &= f \otimes g + f \otimes h & 
% (f+g) \otimes h &= f \otimes h + g \otimes h,
% %\item A set $ob(\mathbb{C})$ of objects;
% %\item For every pair $A,B \in ob(\mathbb{C})$, a hom-semigroup $Hom(A,B)$;
% %\item An element $\textsf{id}_{A,B} \in Hom(A,B)$; 
% \end{align*}
% Note,  we take $(;\!)$ and $\otimes$ to bind more tightly than $+$.
% A \textit{semilattice enriched strict SMC functor} $F: (\mathbb{C}, \otimes, 1,+) \to (\mathbb{D}, \otimes, 1,+)$ is given by a strict SMC functor on the underlying SMCs which additionally satisfies $F(f+g) = F(f)+F(g)$.
% All definitions extend to \textit{PROP} when the manipulated categories are PROPs rather than SMCs. 
% We denote the \textit{freely enriched PROP (SMT)} over a monoidal signature $\Sigma$ as $\textbf{PROP}^+(\Sigma)$ ($\textbf{SMT}^+(\Sigma, \mathcal{E})$,  respectively). 
% \end{definition}
\begin{definition}
We define $\catname{S}(\Sigma)^{+}$ and $\catname{S}(\Sigma, \mathcal{E})^{+}$ to be semilattice-enriched SMCs obtained as $\mathcal{F}(\catname{S}(\Sigma))$ and $\mathcal{F}(\catname{S}(\Sigma, \mathcal{E}))$ respectively.
Both of these categories are in $\catname{PROP}^{+}$ that is obtained from $\catname{PROP}$ via the same adjunction.
\end{definition}
Equivalently, they can be constructed from taking (typed) $\Sigma^+$-terms  -- namely, those freely constructed from generators $c \in \Sigma$, $\textsf{id}_1$, $\sym_{1,1}$, $(;\!)$ and $\otimes$ (and $f+g$, respectively) -- quotiented by the axioms of an enriched symmetric monoidal category (and the least congruence including $\mathcal{E}$ in the case of $\catname{S}(\Sigma, \mathcal{E})^{+}$).
% The free category $\textbf{PROP}^+(\Sigma)$ can be syntactically constructed from taking (typed) $\Sigma^+$-terms  -- namely, those freely constructed from generators $c \in \Sigma$, $\textsf{id}_1$, $\sym{1,1}$, $(;\!)$ and $\otimes$ (and $f+g$, respectively) -- quotiented by the axioms of an enriched symmetric monoidal category.

To aid reasoning, we introduce a new language of string diagrams for semilattice enriched SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms.  It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting.

@@

\ifdefined \ONECOLUMN
\begin{figure}
	\[  
		\scalebox{0.6}{
		\tikzfig{../figures/categorical-semantics/egraph-strings-equations}
		}
	\]
	\caption{Equations for a  semilattice enriched symmetric monoidal category}
	\label{fig:string-equations}
	\end{figure}
\else
\begin{figure*}
\[  
    \scalebox{0.75}{
	\tikzfig{../figures/categorical-semantics/egraph-strings-equations}
    }
\]
\caption{Equations for a  semilattice enriched symmetric monoidal category}
\label{fig:string-equations}
\end{figure*}
\fi

>>>>

\ifdefined \ONECOLUMN
\begin{figure}
	\[
		\scalebox{0.55}{
		\tikzfig{../figures/categorical-semantics/egraph-translation}
		}
	\]
	\caption{Example translation of acyclic e-graphs into string diagrams for semilattice-enriched Cartesian categories. }
	\label{fig:e-graph-example}
	\end{figure}
\else
\begin{figure*}
\[
	\hspace{1.3cm}
    \scalebox{0.65}{
    \tikzfig{../figures/categorical-semantics/egraph-translation}
    }
\]
\caption{Example translation of acyclic e-graphs into string diagrams for semilattice-enriched Cartesian categories. }
\label{fig:e-graph-example}
\end{figure*}
\fi

@@

In order to \textit{implement} generalised e-graphs, rewriting must be performed on concrete representations of the corresponding string diagrams.  
String diagrams can be considered equivalently as either topological objects (\textit{i.e.}, taken modulo ``connectivity'') or as a 2-dimensional syntax quotiented by the equations of a \textit{symmetric monoidal category (SMC)}.
For example, we have the following equivalences between diagrams:
%\[(f_1 \otimes f_2) ; (g_1 \otimes g_2) = (f_1 ; g_1) \otimes (f_2 ; g_2)\] 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/interchange}
	}
\]
The latter representation makes string diagrams unamenable to efficient implementation due to the difficulty of calculating the quotient.
\emph{Hypergraphs} are a combinatorial representation designed to address this issue~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.
% A large body of research work shows how this issue can be solved by taking a different approach, representing string diagrams as hypergraphs~\cite{bonchi_string_2022-1,bonchi_string_2022-2,bonchi_string_2022}.  
Here, the generators $c_i$ become hyper-edges,  and wires become vertices.  
Thus, the expected quotient becomes simply hypergraph isomorphism. 
With appropriate restrictions on the form these hypergraphs can take, this approach can be used to \textit{characterise} the free symmetric monoidal category generated by some~$c_i$.

We are interested not only in the free SMC over a set of generators, but also in \textit{(symmetric) monoidal theories} with extra equations such as, for instance, the rewrite rules (b)--(d) of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus rewrites of their hypergraph representations.  Because the generating equations can be applied in any context, we are led to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ within a hypergraph and replace it with (a sub-hypergraph corresponding to) $r$.
The standard methodology to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting} \cite{dpo, bonchi_string_2022}, is still applicable.
However, these concepts must be now generalised from hypergraphs to \textit{e-hypergraphs} --- hypergraphs with two additional relations denoting the hierarchical structure introduced by so-called \textit{e-boxes}: the generator for semilattice enrichment, and the separation of the components of each e-box.
Giving the appropriate definitions turns out to be a technical matter of some complexity, and constitutes the main body of the paper.

\subsection{Soundness and Completeness}
The main technical result of the paper is a proof of soundness and completeness of this representation with respect to morphisms of a semilattice-enriched symmetric monoidal categories and their respective rewriting.
This extends the results of \cite{bonchi_string_2022-2} for plain SMCs. 
While the structural equalities of SMCs are factored out in the representation,  those arising from the enrichment (see the equations of Figure \ref{fig:string-equations}) are not, and should not be. 
They represent the un/sharing of subdiagrams with respect to the e-box structure,  which is precisely what allows for the compact representation of equivalence classes.  
Instead, we consider DPO-rewrites implementing both the structural equalities for enrichment (which involve the e-box structure) and the equalities arising from the generating monoidal theory (which do not).

Precisely,  our soundness and completeness result is the statement that morphisms in an appropriate free semilattice-enriched SMC are equal  if and only if there exists a sequence of DPO-rewrites --- each induced by a structural equality or the monoidal theory --- between their combinatorial representations. 
In the particular case of a monoidal signature including natural copy and delete maps,  we thus recover a sound and complete representation of semilattice-enriched Cartesian categories,  with the translation described earlier in Figure~\ref{fig:e-graph-to-string} justifying our claim of developing a mathematical theory of e-graphs.

%\subsection{E-matching and E-rewriting}
%
%Having defined e-hypergraphs and shown them to correctly model the rewriting of string diagrams for semilattice-enriched SMCs, we note that the naive implementation of DPO-rewriting is inefficient: to find a redex within an e-hypergraph involves finding a structurally equivalent e-hypergraph which contains the redex as a subgraph. In general, this involves unsharing the e-box structure before a redex becomes available.  Thus, we define a notion of \textit{e-matching} for e-hypergraphs: that is, to find redexes working modulo the e-box structure. In particular, we wish to locate the smallest subgraph $G' \subseteq G$ "containing" (in an appropriate sense) the redex $L$. Given this,  \textit{e-rewriting} $L \to R$, intended to achieve equality saturation, is simple to define due to its non-destructive nature: we rewrite $G' \to G' + R$ in $G$.  
%\begin{figure}
%\[
%	\tikzfig{combinatorial_semantics/example-rewriting}
%\]
%\caption{Example application of e-rewriting for rewrite $(x*y)/z \to x*(y/z)$: find an appropriate minimal sub-e-hypergraph containing the redex $(x*y)/z$ and non-destructively add the reduct $x*(y/z)$.  REDO THIS}
%\end{figure}

@@

% E-graphs
Although e-graphs were first developed in the 80s \cite{nelson1980techniques}, there has recently been an explosion of interest in them for the purpose of building program optimisers and synthesisers, especially due to recent work on the equality saturation technique \cite{10.1145/1594834.1480915, griggio_proceedings_2022, EggPaper,flatt_small_2022}.  This includes interest in various extensions of the e-graph formalism, for example to account for conditional rewriting \cite{singher2023colored},  to reason about rewriting for the $\lambda$-calculus \cite{koehler2022sketchguided},  and to combine techniques of e-graphs and database queries (``relational e-matching'') \cite{zhang_relational_2022}. 
We are hopeful that our novel perspective on e-graphs can be extended to give uniform foundations to these investigations.  
Some further avenues for investigation, including potential applications, are given in the \nameref{sec:conclusion}.

@@

\section{Categorical Semantics of E-Graphs}
In this section,  we introduce the required preliminaries on semilattice-enriched symmetric monoidal categories generated by monoidal theories,  and the string diagrammatic formalism we will use to represent them.

Given a category $\mathbb{C}$  with objects $A,B \in \mathbb{C}$ we denote by $\mathbb{C}(A,B)$ the corresponding hom-set.  
We write the identity morphism on $A$ as $\id_A$,  or simply $A$.  
We commonly write $f;g$ for composition in diagrammatic order.  
Composition in the usual order is written $g \circ f$.  
We denote the tensor product of an SMC $\mathbb{C}$ by $\otimes$,  its unit by $I$, and its symmetry natural transformation as $\sym$ \cite{maclane}.  
We adopt the convention that $\otimes$ binds more tightly than $(;\!)$.  
We elide all associativity and unit isomorphisms associated with monoidal categories,  and often omit subscripts on identities and natural transformations where it can be inferred.  
We denote by $\catname{SLat}$ the closed monoidal category of semilattices which is defined in section~\ref{sec:appendix:1} of the Appendix.

@@

It will often be convenient to give presentations of SMTs in terms of string diagrams.
\begin{example}
The SMT of \textit{commutative comonoids} is given by the following generators, ${\Delta, !}$, depicted in string diagrammatic form:
\[
	\scalebox{0.8}{
  	 \tikzfig{../figures/categorical-semantics/Cartesian-equipment}
	}
\]
together with the following associativity, commutativity and unitality equations, given in terms of equations of string diagrams. 
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-theory}	
	}
\]
A Cartesian SMT is given by a set of generators $\Sigma_C$ which additionally contains the generators of commutative comonoids, together with the equations of the SMT of commutative comonoids, and the following additional naturality of copy-delete equations, for every $c \in \Sigma_C$
\[
	\scalebox{0.6}{
	\tikzfig{../figures/categorical-semantics/Cartesian-naturality}
	}
\]
By Fox's Theorem~\cite{fox},  $\textbf{S}(\Sigma_C, \mathcal{E}_C)$ is equivalent to the free Cartesian category over $\Sigma_C$. 
\end{example}

@@

We will use enrichment to axiomatize the e-box (e-class) structure as the ``join" of two morphisms $f,g: A \to B$ as $f + g: A \to B$. 
More precisely our hom-sets will be given the structure of a semilattice.
% Note, in particular,  that we do not require semilattices to have a unit for $+$. 
% This is important, since, in any commutative monoid enriched category,  if there exists a categorical product, then it must be a categorical biproduct \cite{maclane}---a degeneracy we wish to avoid.  
Generally,  we may define a category \textit{enriched} in any other (closed) monoidal category $M$ (\textit{e.g.},  $\catname{SLat}$).
In this case,  hom-sets are generalised to hom-objects of $M$ (\textit{e.g.},  a semilattice) and composition 
\[
	\circ: Hom(B,C) \otimes Hom(A,B) \to Hom(A,C)
\]
is defined as an $M$-morphism which satisfies certain axioms.
Composition being an $M$-morphism implies it respects the additional structure on hom-sets.
%  In other words,  composition must respect the additional structure on hom-sets.  
% In our case,  the category we wish to enrich is also monoidal,  so we additionally ask that the monoidal structure also respects this structure.  Technically, this amounts to asking that the tensor product is an enriched functor. 
We omit further details of the general case of enrichment here, and instead work with the concrete definition required for our application, where $M = \catname{SLat}$.

\begin{definition}[Semilattice-enriched category]
A semilattice-enriched category $\mathbb{C}$ is defined by the following data
\begin{itemize}
	\item a set of objects $\mathbb{C}$
	\item for every pair of objects $A,B \in \mathbb{C}$ --- a hom-object (or, hom-semilattice) $\mathbb{C}(A,B) \in \catname{SLat}$
	\item for every triple of objects $A,B,C \in \mathbb{C}$ --- a composition morphism
	\[
		\circ: \mathbb{C}(B,C) \otimes \mathbb{C}(A,B) \to \mathbb{C}(A,C)
	\]
	\item for every object $A \in C$ --- a unit morphism
	\[
	u_{A} : I \to \mathbb{C}(A,A), \text{ where $I$ is the monoidal unit for $\catname{SLat}$}
	\]
\end{itemize}
such that particular coherence diagrams commute~\cite{Borceux_1994}.
We will also call such categories $\catname{SLat}$-categories.
\end{definition}

@@

Note, in particular,  that we do not require semilattices to have a unit for $+$, \textit{i.e.,} we work with \textit{unbounded} semilattices.
% We also will not utilize the ordering that is induced by the operator of a semilattice in any meaningful way, so a semilattice can be thought of as a commutative idempotent semigroup throughout the paper.
% We stick to the term semilattice as it is shorter.

\begin{definition}[$\catname{SLat}$-functor]
	Let $\mathbb{C}$ and $\mathbb{D}$ be two $\catname{SLat}$-categories.
	An $\catname{SLat}$-functor $F : \mathbb{C} \to \mathbb{D}$ is defined by the following data.
	\begin{enumerate}
	  \item A mapping $F : \mathbb{C} \to \mathbb{D}$
	  \item An object-indexed family of morphisms in $\catname{SLat}$ $F_{A,B} : \mathbb{C}(A,B) \to \mathbb{D}(FA,FB)$
	\end{enumerate}
	such that certain coherence diagrams commute~\cite{Borceux_1994}.
	$F$ being a morphism in $\catname{SLat}$ has a property $F(f + g) = F(f) + F(g)$.
	An $\catname{SLat}$-functor $F$ defines an equivalence between two $\catname{SLat}$-categories if it is \textit{full}, \textit{faithful}, and \textit{essentially surjective}~\cite{Kelly2022BASICCO}.
\end{definition}

\begin{definition}[$\catname{SLat}$-SMC]\label{def:enriched-prop}
A \textit{semilattice-enriched strict SMC}  $\mathbb{C}$ is given by an $\catname{SLat}$-category $\mathbb{C}$ as above that additionally has
\begin{itemize}
\item a unit object $I_{C} \in \mathbb{C}$
\item for every pair of objects $A,B \in \mathbb{C}$ --- an object $A \otimes B \in \mathbb{C}$
\item for all $A,B,C,D$ --- a tensor morphism $ - \otimes_{C} - : \mathbb{C}(A,C) \otimes \mathbb{C}(B,D) \to \mathbb{C}(A \otimes B, C \otimes D)$
\end{itemize}
such that particular coherence diagrams commute~\cite{enriched_monoidal}.
\end{definition}
The latter morphism being in $\catname{SLat}$ implies the following equations that $\otimes$ satisfy
\[
f \otimes (g+h) = f \otimes g + f \otimes h \qquad (f+g) \otimes h = f \otimes h + g \otimes h
\]

@@

\begin{proposition}(A specialized case of Proposition 6.4.3~\cite{Borceux_1994}. See also Appendix section~\ref{sec:appendix:1})
There is a 2-adjunction 
% https://q.uiver.app/#q=WzAsMixbMCwwLCJcXGNhdG5hbWV7U0xhdHR9LVxcY2F0bmFtZXtDYXR9Il0sWzIsMCwiXFxjYXRuYW1le0NhdH0iXSxbMSwwLCJGIiwyLHsiY3VydmUiOjR9XSxbMCwxLCJHIiwyLHsiY3VydmUiOjR9XSxbMiwzLCIiLDAseyJsZXZlbCI6MSwic3R5bGUiOnsibmFtZSI6ImFkanVuY3Rpb24ifX1dXQ==
\[\begin{tikzcd}
	\catname{SLatt}\text{--}\catname{Cat}\arrow[rr, "\mathcal{U}", bend left] & \hspace{-1em}\top & \catname{Cat} \arrow[ll, "\mathcal{F}", bend left]
	\end{tikzcd}
\]
that is induced by a usual free-forgetful adjunction

@@

In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice-enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
% SMC $(\mathbb{C}, \otimes, 1,+)$ where every hom-set additionally has the structure of a semilattice $(\mathbb{C}(A,B), +)$ -- that is,  a set equipped with an associative,  commutative,  and idempotent operation -- which respects the composition and tensor product in the following ways,  for all appropriately typed morphisms.
% \begin{align*}
% f ; (g+h) &= f;g + f;h &
% (f+g) ; h &= f;h + g;h \\
% f \otimes (g+h) &= f \otimes g + f \otimes h & 
% (f+g) \otimes h &= f \otimes h + g \otimes h,
% %\item A set $ob(\mathbb{C})$ of objects;
% %\item For every pair $A,B \in ob(\mathbb{C})$, a hom-semigroup $Hom(A,B)$;
% %\item An element $\textsf{id}_{A,B} \in Hom(A,B)$; 
% \end{align*}
% Note,  we take $(;\!)$ and $\otimes$ to bind more tightly than $+$.
% A \textit{semilattice-enriched strict SMC functor} $F: (\mathbb{C}, \otimes, 1,+) \to (\mathbb{D}, \otimes, 1,+)$ is given by a strict SMC functor on the underlying SMCs which additionally satisfies $F(f+g) = F(f)+F(g)$.
% All definitions extend to \textit{PROP} when the manipulated categories are PROPs rather than SMCs. 
% We denote the \textit{freely enriched PROP (SMT)} over a monoidal signature $\Sigma$ as $\textbf{PROP}^+(\Sigma)$ ($\textbf{SMT}^+(\Sigma, \mathcal{E})$,  respectively). 
% \end{definition}
\begin{definition}
We define $\catname{S}(\Sigma)^{+}$ and $\catname{S}(\Sigma, \mathcal{E})^{+}$ to be semilattice-enriched SMCs obtained as $\mathcal{F}(\catname{S}(\Sigma))$ and $\mathcal{F}(\catname{S}(\Sigma, \mathcal{E}))$ respectively.
Both of these $\catname{SLat}$-categories are in $\catname{PROP}^{+}$ which we define to be the image of $\catname{PROP}$ via the same adjunction.
\end{definition}
Equivalently, they can be constructed from taking (typed) $\Sigma^+$-terms  -- namely, those freely constructed from generators $c \in \Sigma$, $\textsf{id}_1$, $\sym_{1,1}$, $(;\!)$ and $\otimes$ (and $f+g$, respectively) -- quotiented by the axioms of an enriched symmetric monoidal category (and the least congruence including $\mathcal{E}$ in the case of $\catname{S}(\Sigma, \mathcal{E})^{+}$).
% The free category $\textbf{PROP}^+(\Sigma)$ can be syntactically constructed from taking (typed) $\Sigma^+$-terms  -- namely, those freely constructed from generators $c \in \Sigma$, $\textsf{id}_1$, $\sym{1,1}$, $(;\!)$ and $\otimes$ (and $f+g$, respectively) -- quotiented by the axioms of an enriched symmetric monoidal category.

To aid reasoning, we introduce a new language of string diagrams for $\catname{SLat}$-SMCs, using a hierarchical ``box'' structure to capture the join operation on morphisms.  It was already discussed in the introduction how this is  used in the translation of equivalence classes from the e-graph to the string diagrammatic setting.

@@

\ifdefined \ONECOLUMN
\begin{figure}
	\[  
		\scalebox{0.6}{
		\tikzfig{../figures/categorical-semantics/egraph-strings-equations}
		}
	\]
	\caption{Equations for a  semilattice-enriched symmetric monoidal category}
	\label{fig:string-equations}
	\end{figure}
\else
\begin{figure*}
\[  
    \scalebox{0.75}{
	\tikzfig{../figures/categorical-semantics/egraph-strings-equations}
    }
\]
\caption{Equations for a  semilattice-enriched symmetric monoidal category}
\label{fig:string-equations}
\end{figure*}
\fi
